How to Avoid OOTA Without Really Trying

### 2024-10-03 paper/ootavacuous.tex
General wordsmithing in multiple places
Various small changes to the text to make it smoother or more
explicit, or to remove unimportant information.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

A computer may execute a thread's instructions out of order.
The architecture specifies the extent to which this may happen, and it
also specifies circumstances under which instructions
must be executed in order.
Nevertheless, we will consider an execution to be determined by the
values obtained by its loads.
As with abstract executions, if two hardware executions of the same
thread on the same type of computer agree on the values obtained by
the load instructions during their first $N$ steps then they will
agree in every respect during those steps, although they may diverge
later.

@@

Invented atomic loads are problematic:
\begin{quote}
\begin{verbatim}
int r0 = x;
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
For values of \co{x} small enough to avoid overflow, the abstract machine is
guaranteed to produce a perfect square in \co{r1}.

But if the compiler is permitted to invent atomic loads then the compiler
might transform this code as follows:
\begin{quote}
\begin{verbatim}
int r0 = x;
int invented = x;
int r1 = r0 * r0 + 2 * invented + 1;
\end{verbatim}
\end{quote}
If the load into \co{r0} happened when the value of \co{x}
was zero and the load into \co{invented} happened when the value of
\co{x} was 10, then the value of \co{r1} will be 21, which is not
a perfect square.
This fails to satisfy the abstract machine's guarantee
and therefore is an invalid transformation.

Inventing atomic loads can also result in OOTA-like behavior.
For example, consider this code, in which the final values of \co{r1}
and~\co{r2} are observable:
\begin{quote}
\begin{verbatim}
   int r1 = (x != 0);
   int r2 = (y != 0);
   z = (r1 == r2);
\end{verbatim}
\end{quote}
It is clear that the store to \co{z} semantically depends on the load
from \co{y}, because the value of \co{z} will change whenever \co{y}
changes between zero and nonzero (all else being equal).

@@

Dependencies between machine instructions correspond to the flow of
information within a CPU.
Each instruction takes a set of inputs and provides a set of outputs,
some of which flow to inputs of later instructions.
The inputs determine what an instruction will do.

@@

No implementation in which semantic dependencies satisfy the Fundamental
Property has a nontrivial OOTA cycle.
To see this, suppose that abstract execution $A$ having an OOTA cycle
is realized by hardware execution $H$.
This means there are atomic stores $W_i$ in $A$,
semantically depending on atomic loads
$\{R_{i,j}\}$ where each of the loads reads from one of the $W_k$
stores in a different thread.
Let $W_i'$ and $R_{i,j}'$ be the hardware instructions these accesses
map to in $H$, if they aren't omitted.
Assuming that the stores are not all omitted,
one of the $W_i'$ instructions, let's say $W_0'$, must commit first.
By the Fundamental Property, one of the loads that $W_0$ depends on,
let's say $R_{0,0}$, is not omitted and $R_{0,0}'$ is ordered before $W_0'$.
But now we have a contradiction:
(1)~$W_0'$ commits after $R_{0,0}'$;
(2)~$R_{0,0}'$ commits after the store instruction $W_k'$ it reads from;
and finally
(3)~$W_k'$ commits no earlier than $W_0'$.

@@

A corollary of this result is that if an implementation's compiler
either
(1)~uses per-thread analysis and treats atomic objects as quasi
volatile, or
(2)~uses global analysis and treats atomic objects as volatile,
then programs produced by that implementation will never exhibit OOTA.
Thus the implementation will automatically comply with full C++, even
though it may be been designed only to comply with loose C++.

@@

One drawback of our definition of semantic dependency is that it is only
relative to a specific compiler or implementation.
This may strike some people as wishy-washy and avoiding the real
problem, in that a given piece of code should either contain or not
contain a semantic dependency, independent of the implementation used to
run it.

>>>>

A CPU may execute a thread's instructions out of order.
The architecture specifies the extent to which this may happen, and it
also specifies circumstances under which instructions
must be executed in order.
Nevertheless, we will consider an execution to be determined by the
values obtained by its loads.
As with abstract executions, if two hardware executions of the same
thread on the same type of computer agree on the values obtained by
the load instructions during their first $N$ steps then they will
agree in every respect during those steps, although they may diverge
later.

@@

Invented atomic loads are problematic.
Consider this code:
\begin{quote}
\begin{verbatim}
int r0 = x;
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
For values of \co{x} small enough to avoid overflow, the abstract machine is
guaranteed to produce a perfect square in \co{r1}.

But if the compiler is permitted to invent atomic loads then the compiler
might transform this code as follows:
\begin{quote}
\begin{verbatim}
int r0 = x;
int rinvented = x;
int r1 = r0 * r0 + 2 * rinvented + 1;
\end{verbatim}
\end{quote}
If the load into \co{r0} happened when the value of \co{x}
was zero and the load into \co{rinvented} happened when the value of
\co{x} was 10, then the value of \co{r1} would be 21, which is not
a perfect square.
This fails to satisfy the abstract machine's guarantee
and therefore is an invalid transformation.

Inventing atomic loads can also result in OOTA-like behavior.
For example, consider this code:
\begin{quote}
\begin{verbatim}
   int r1 = (x != 0);
   int r2 = (y != 0);
   z = (r1 == r2);
\end{verbatim}
\end{quote}
It is clear that the store to \co{z} semantically depends on the load
from \co{y}, because the value of \co{z} will change whenever \co{y}
changes between zero and nonzero (all else being equal).

@@

Dependencies between machine instructions are determined by the flow of
information within a CPU.
Each instruction takes a set of inputs and provides a set of outputs,
some of which flow to inputs of later instructions.
The inputs determine what an instruction will do.

@@

No implementation in which semantic dependencies satisfy the Fundamental
Property can realize a nontrivial OOTA cycle.
To see this, suppose that abstract execution $A$ having an OOTA cycle
is realized by hardware execution $H$.
This means there are atomic stores $W_i$ in $A$,
semantically depending on atomic loads
$\{R_{i,j}\}$ where each of the loads reads from one of the $W_k$
stores in a different thread.
Let $W_i'$ and $R_{i,j}'$ be the hardware instructions these accesses
map to in $H$, if they aren't omitted.
Assuming that the stores are not all omitted,
one of the $W_i'$ instructions, let's say $W_0'$, must commit first.
By the Fundamental Property, one of the loads that $W_0$ depends on,
let's say $R_{0,0}$, is not omitted and $R_{0,0}'$ is ordered before $W_0'$.
But now we have a contradiction:
(1)~$W_0'$ commits after $R_{0,0}'$;
(2)~$R_{0,0}'$ commits after the store instruction $W_k'$ it reads from;
and finally
(3)~$W_k'$ commits no earlier than $W_0'$.

@@

A corollary of this result is that if an implementation's compiler
either
(1)~uses per-thread analysis and treats atomic objects as quasi
volatile, or
(2)~uses global analysis and treats atomic objects as volatile,
then programs produced by that implementation will never exhibit OOTA.
Thus the implementation will automatically comply with full C++, even
though it may have been designed only to comply with loose C++.

@@

One drawback of our definition of semantic dependency is that it is only
relative to a specific compiler or implementation.
This may strike some readers as wishy-washy and avoiding the real
problem, in that a given piece of code should either contain or not
contain a semantic dependency, independent of the implementation used to
run it.
